Fast Sorting on Modern Computers

### 2005-08-30 docs/tech_report/heap.tex
fixed heapsort. rewrote a bit of fututre work. changed a H_N to h_N-1

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
1.59 mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches harmonics}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions (about 36\%) due to the combination of the shallower heap
and the larger fan-out.

@@

The branch prediction behaves differently to the expected results, as the
prediction ratio gets worse as the fanout gets larger. The reason we predicted
differently was due to selection sort's $O(H_N)$ mispredictions; however, this
is over the entire length of the array, and it is not accurate to extrapolate
these results over a sequence only seven keys long.

@@

It should be possible to more accurately model and therefore predict the branch
prediction ratio when the array is short. Our results show us what we expect the
answer to be, but have not generated a formula to predict it.

>>>>

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
1.59 mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches harmonics}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions due to the combination of the shallower heap and the larger
fan-out.

@@

The branch prediction behaves differently to the expected results, as the
prediction ratio gets worse as the fanout gets larger. The reason we predicted
differently was due to selection sort's $O(H_N-1)$ mispredictions; however, this
is over the entire length of the array, and it is not accurate to extrapolate
these results over a sequence only seven keys long.

@@

The formula we use to explain selection sort's branch prediction preformance
breaks down for short arrays, as in our 8-heap. It should be possible to more
accurately model and therefore predict the branch prediction ratio when the
array is short. Our results show us what we expect the answer to be, but we have
not attempted to create a formula to predict it.
